

---
title: "musicassessr melodic features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{musicassessr melodic features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE, warning = FALSE, message = FALSE}

library(dplyr)

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

```

The melodic features used in `musicassessr` mainly come from the *FANTASTIC* library (Müllensiefen, 2009), except where noted.

```{r}

itembankr::get_melodic_feature_table(return_kable = FALSE) %>% 
  mutate(Feature = case_when(Feature == "target_melody_length" ~ "N", TRUE ~ Feature)) %>% 
  knitr::kable(escape = FALSE)

```


# References


Beaty, R. E., Frieler, K., Norgaard, M., Merseal, H. M., MacDonald, M. C., & Weiss, D. J. (2021). Expert musical improvisations contain sequencing biases seen in language production. Journal of Experimental Psychology. https://doi.org/10.1037/xge0001107


Harrison, P. M. C., Bianco, R., Chait, M., & Pearce, M. T. (2020). PPM-Decay: A computational model of auditory prediction with memory decay. PLOS Computational Biology, 16(11), e1008304. https://doi.org/10.1371/journal.pcbi.1008304

Müllensiefen, D. (2009). FANTASTIC: Feature ANalysis Technology Accessing STatistics (In a Corpus; Technical report). 37.

